\chapter{Background}
\label{ch:background}

In this chapter we will provide the necessary background and graph theory concepts needed for the rest of the thesis.
We will also provide a brief overview of classical algorithms for the solution of the Shortest Path Problem
(e.g. Dijkstra's\cite{dijkstraNoteTwoProblems1959}, A*\cite{hartFormalBasisHeuristic1968b}, ALT\cite{goldbergComputingShortestPath2005}, Contraction Hierarchies\cite{geisbergerContractionHierarchiesFaster2008}),
as well as newer algorithms developed in the area of \textit{Algorithms with Predictions} (or \textit{Learning Augmented Algorithms}).

\section{Graph Theory Basics}
We will start by defining some basic graph theory concepts that will be used throughout the thesis.
For the purposes of this thesis, we will use the standard notations as described in Cormen et al.\cite{cormenIntroductionAlgorithms2009}
\footnote{Definitions and notation follow Cormen et al.~\cite{cormenIntroductionAlgorithms2009}, specifically Appendix B.4 (Graph Theory)}

%definitions below are based on Cormen et al. Introduction to Algorithms (2009). if i used a wrong definition please let me know :) 
A \textit{directed graph} $G$ is defined as an ordered pair $G=(V, E)$ where $V$ is a non-empty set of vertices (or nodes/points) and $E$ is a set of edges (or arcs/lines) that connect pairs of vertices.
We will denote the number of vertices as $n = |V|$ and the number of edges as $m = |E|$.

A \textit{weighted graph} is a graph $G$ where each edge has a numerical value or \textit{weight} (or \textit{cost}) that represents the cost of traversing that edge.
For a given edge $e \in E$, we will denote its weight as $w(e)$.
We define a weight function $w: E \rightarrow \mathbb{R}_{\geq 0}$
that assigns a non-negative
\footnote{The assumption of non-negative weights is necessary for the correctness of Dijkstra's algorithm, which we will revisit on Section 2.3.}
real number to each edge in the graph.

A \textit{path} $P$ is defined as a sequence of vertices $P = (v_1, v_2, ..., v_k)$ from a source vertex $s$ to a target vertex $t$ 
such that for $v_1 = s$, $v_k = t$, and $(v_i, v_{i+1}) \in E$ for all $1 \leq i < k$.
The \textit{cost} (or \textit{length}) of a path $P$ is defined as the sum of the weights of the edges in the path, i.e.

\begin{equation}
    w(P) = \sum_{i=1}^{k-1} w(v_i, v_{i+1})
\end{equation}

We denote \textit{Shortest Path Distance} from $s$ to $t$ as $\delta(s, t)$, which is the minimum cost from $s$ to $t$ compared to all possible paths from $s$ to $t$. 
If no path exists from $s$ to $t$, we define $\delta(s, t) = \infty$.

\section{The Shortest Path Problem}
The Shortest Path Problem is one of the most studied problems in graph theory.
The problem focuses on finding the minimum cost from a starting vertex $s$ to a every other vertex $v \in V$
In other words finding the \textit{Shortest Path Distance} from $s$ to all $v \in V$ similarly to how we previously mentioned ($\delta(s,v)$), and construct the corresponding paths that achieve this cost.

This problem is used in technologies we use daily such as in GPS navigation, network routing, logistic networks and generally in any path finding need.
While many simple cases can be solved easily just by intuition, in today's era of Big Data we require the use of algorithms (like \textsc{DIJKSTRA}) to solve more complex and interconnected cases.

\subsection{SPP Variations}
There are many variations of the Shortest Path Problem that were analyzed and studied in order to satisfy the different needs of each use case.

The \textit{All-Pair Shortest Path} (APSP) Problem for example searches for the shortest path between every possible pair of nodes ($u , v$)in a graph.
A famous algorithm for the solution of this problem was designed by Robert W. Floyd\cite{floydAlgorithm97Shortest1962} in 1962 with a time complexity of $O(n^3)$.
Further more, in the cases of sparse graphs an algorithm was designed by Donald B. Johnson\cite{johnsonEfficientAlgorithmsShortest} in 1977 further reducing the time complexity to $O(n^2\log n+nm)$
(Although the algorithm performs the same of slightly worse in the case of dense graphs).

The \textit{Single-Destination Shortest Path} (SDSP) Problem searches for the shortest path from every node $v \in V$ to a single target node $t$. 
In this case one can solve the problem simply by reversing the order of the graph edges and running Dijkstra's algorithm\cite{dijkstraNoteTwoProblems1959} (which we will develop on Section 2.3).
It is important to note that in the case of an unordered graph the solution for the Single-Destination Shortest Path is the same as the solution of the Single-Source Shortest Path.

Lastly, another notable variation to mention is the \textit{k-Shortest Path} (kSP) Problem which instead of searching for only the shortest possible path, it searches for the k-shortest possible paths. 
This variation is commonly used in navigation systems offering user an alternative route to take besides the shortest one possible. 
A solution for this problem was designed by Jin Y. Yen\cite{yenFindingShortestLoopless1971a} in 1971 by executing Dijkstra's algorithms $O(n)$ times for each of the k shortest paths with a time complexity of $O(kn(E+n\log n))$.

\subsection{Single-Source Many-Targets Shortest Path}
In this thesis, we focus specifically on the \textit{Single-Source Many-Targets Shortest Path} (SSMTSP) problem, which is the version of the Shortest Path Problem addressed by the algorithm of Feijen and Schäfer~\cite{feijenDijkstrasAlgorithmPredictions2023}.

SSMTSP is a variation of the Single-Source Shortest Path (SSSP) problem, where instead of finding the shortest path from a source node to all other nodes in the graph, we search to find the shortest paths from a single source node to a specific set of target nodes.
More specifically:

Given a directed, weighted graph $G=(V, E)$, with a weight function 
\footnote{As mentioned before the assumption of $\mathbb{R}_{\geq 0}$ is necessary for Dijkstra's algorithm}
$w: E \rightarrow \mathbb{R}_{\geq 0}$
, a source node $s$, and a set of target nodes $T \subseteq V$, the goal is to find the shortest paths $\delta(s,t)$ from $s$ to all $t \in T$.

This specific version of the problem is used in real-world scenarios such as in Points of Interest (POI) search in navigation systems, where a user in a specific location (source node) may want to find multiples options (target nodes) for a specific destination (e.g. gas stations, restaurants, etc.) and not just the closest one.
It is also used as a subroutine in logistics and specifically in the Last-Mile Delivery problem, where a distribution center (source node) needs to deliver goods to multiple customers
\footnote{Usually the delivery routine is calculated with a variation of the \textit{Traveling Salesperson Problem} (TSP), but often the first step is to calculate the shortests paths by solving the SSMTSP problem}
(target nodes) in the most efficient way possible.
Furthermore, it is used in Content Delivery Networks (CDNs) where a server (source node) needs to deliver content to multiple users (target nodes) in the most efficient way possible. 

\section{Dijkstra's Algorithm}
The most famous and widely used algorithm for the SSSP Problem and specifically for our case the SSMTSP Problem is the \textsc{DIJKSTRA} Algorithm\cite{dijkstraNoteTwoProblems1959} which was proposed by Edsger W. Dijkstra in 1959. 

\textsc{DIJKSTRA} works only in cases where the edges have a non-negative weight ($w(e) \ge 0$ for each $e \in E$). 

\subsection{Algorithmic Analysis}
\textsc{DIJKSTRA}\footnote{Definitions and notation follow Cormen et al.~\cite[Ch.~24.3]{cormenIntroductionAlgorithms2009}, specifically Chapter 24.3 (Dijkstra's Algorithm)} works with a set S of visited (or settled) nodes. 
At the start $S$ is an empty set, the distance to the source node $s$ is 0 and the distance to the rest of the nodes in the graph is $\infty$.
We repeat a process of taking the unvisited (or unsettled) node $u \in V-S$ with the minimum distance, adding $u$ to $S$, and \textit{relaxing} all edges leaving $u$. 

For a given edge $(u, v) \in E$ with weight $w(u, v)$, if the current known distance to $v$ is greater than the distance to $u$ plus the weight of the edge, 
the algorithm updates (\textit{relaxes}) the distance to $v$:

\begin{equation}
\text{If } dist(v) > dist(u) + w(u, v) \text{, then update } dist(v) = dist(u) + w(u, v)
\end{equation}

The exact code implementation of \textsc{DIJKSTRA} we will use is as follows:

\begin{verbatim}
//TODO: ADD LOCAL DIJKSTRA CODE IMPLEMENTATION
\end{verbatim}

\textsc{DIJKSTRA} is a greedy algorithm\footnote{"Greedy" algorithms work by making the best choice in any specific step/iteration (local optimal)} 
since it selects the "lightest" vertex thus selecting a local optimal choice, but unlike most greedy algorithms it guarantees the global optimal result everytime.

As mentioned in the introduction, through the use of \textit{Fibonacci Heaps} as proposed by Fredman and Tarjan\cite{fredmanFibonacciHeapsTheir1987}, 
\textsc{DIJKSTRA} can achieve a time complexity of O($m$ + $n\log n$) (where $n$ and $m$ are the number of nodes and edges respectively).

\subsection{Limitations}
Recently, Duan et al. in their famous paper \textit{Breaking the Sorting Barrier for Directed Single-Source Shortest Paths}\cite{duanBreakingSortingBarrier2025}, 
improved the time complexity of \textsc{DIJKSTRA} to $O(m \log^{2/3}n)$, 
yet \textsc{DIJKSTRA} still has a fundamental flaw, it's working "blind" without knowing whether or not if the path it is exploring will lead towards a target node $t \in T$. 

Specifically in our case, on the SSMTSP problem, even if \textsc{DIJKSTRA} manages to find the shortest path to one of the target nodes 
it still has many paths to explores and thus many more cases where it may explore in completely wrong direction leading to the calculation of irrelevant regions.

This inefficiency highlights the need of algorithms which use external information.
Heuristic algorithms like A*\cite{hartFormalBasisHeuristic1968b} attempt to combat this issue, but even then, we require specific knowledge to calculate the heuristic value $h(n)$, 
on which we have the inability to dynamically re-calculate computationally all the time for multiple targets.

\textsc{DIJKSTRA-PREDICTION} on the other hand, tackles the issue by utilizing machine learning techniques to prune the search space all-together, skipping irrelevant edges, 
while still maintaining an optimal shortest path.